Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-16 paper/paper.tex
Minor updates

Information for each of the features in Table~\ref{tab:features} that are denoted as \colorbox{Gainsboro}{temporal} is only used when it was published in a year \textbf{prior} to the first phase 2 trial for a TD pair. Genetic evidence is a notable exception to this rule, and motivations for that exception are detailed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \ul{2 year} offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of recent outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

@@

Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study, RDG, ranks TD pairs by comparison to a composite score from Open Targets, OTS. This comparison highlights relative risk (RR) as our primary performance indicator along with secondary measures of performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), as discussed more in Section~\ref{sec:metrics}. The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it uses time since the phase 2 transition as a predictive factor in addition to all others. We observe that the use of this information greatly improves standard performance metrics like receiver operating characteristic (ROC) and average precision (AP), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed. In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support. More implications of this and opportunities it may imply are discussed in Section~\ref{sec:discussion}. As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available. They do, however, present a useful performance ceiling towards which future work might build.

@@

Utilizing our predictive models, we systematically identify target-disease pairs that have not yet been explored in clinical trials but exhibit a high likelihood of successful outcomes. This process involved ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.

>>>>

Information for each of the features in Table~\ref{tab:features} that are denoted as \colorbox{Gainsboro}{temporal} is only used when it was published in a year \textbf{prior} to the first phase 2 trial for a TD pair. Genetic evidence is a notable exception to this rule, and motivations for that exception are detailed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a 2 year offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of recent outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

@@

Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study, RDG, ranks TD pairs by comparison to a composite score from Open Targets, OTS. This comparison highlights relative risk (RR) as our primary performance indicator along with secondary measures of performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), as discussed more in Section~\ref{sec:metrics}. We observe improved performance over this benchmark across all metrics, and provide further comparisons between the OTS score and other models in Section~\ref{sec:model_comparison}.

The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it uses time since the phase 2 transition as a predictive factor in addition to all others. We observe that the use of this information greatly improves standard performance metrics like receiver operating characteristic (ROC) and average precision (AP), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed. In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support. More implications of this and opportunities it may imply are discussed in Section~\ref{sec:discussion}. As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available. They do, however, present a useful performance ceiling towards which future work might build.

@@

Utilizing our predictive models, we systematically identify target-disease pairs that have not yet been explored in clinical trials but exhibit a high likelihood of successful outcomes. This process involves ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.
